---
title: "Report"
author: "FH"
date: "August 3, 2020"
output: html_document
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Inroduction


```{r cars}
summary(cars)
```

## Overview

Variable names, summary

```{r pressure, echo=FALSE}
plot(pressure)
```


##Method Section

Explain the method and techniques used, Visualization with insights, data exploration. Use at least two Methods 
You will use the following code to generate your datasets. Develop your algorithm using the edx set. For a final test of your algorithm, predict movie ratings in the validation set (the final hold-out test set) as if they were unknown. RMSE will be used to evaluate how close your predictions are to the true values in the validation set (the final hold-out test set).

##Results Section

Resultes presented and discussed.Model performance. 


##Conclusion Section

Brief summary of the report. It's limitation and future work.

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
