---
title: "Report"
author: "FH"
date: "August 3, 2020"
output: html_document
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Inroduction


```{r cars}
summary(cars)
```

## Overview

Variable names, summary

```{r pressure, echo=FALSE}
plot(pressure)
```


##Method Section
Develop your algorithm using the edx set. For a final test of your algorithm, predict movie ratings in the validation set (the final hold-out test set) as if they were unknown. RMSE will be used to evaluate how close your predictions are to the true values in the validation set (the final hold-out test set).

Important: The validation data (the final hold-out test set) should NOT be used for training your algorithm and should ONLY be used for evaluating the RMSE of your final algorithm. You should split the edx data into separate training and test sets to design and test your algorithm.



Explain the method and techniques used, Visualization with insights, data exploration. Use at least two Methods 
You will use the following code to generate your datasets. Develop your algorithm using the edx set. For a final test of your algorithm, predict movie ratings in the validation set (the final hold-out test set) as if they were unknown. RMSE will be used to evaluate how close your predictions are to the true values in the validation set (the final hold-out test set).

##Results Section

Resultes presented and discussed.Model performance. 


##Conclusion Section

Brief summary of the report. It's limitation and future work.

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
